Lossy compression of DNA sequencing quality data

### 2015-04-13 manuscript/q-compression.tex
Changed -order to -degree, as degree is the more proper term.

We compare the \textsc{mse} versus bits/base-call of the
\textit{Rhodobacter sphaeroides} 2.4.1, \textit{Homo sapiens}
chromosome 14, \textit{Escherichia coli} str. K-12 MG1655, and
\textit{Mus musculus} data sets (Figure \ref{fig:mse_vs_bpbp}). We
only include the fragment libraries for the \textit{Rhodobacter
  sphaeroides} 2.4.1, and \textit{Homo sapiens} chromosome 14 data
sets, but the additional short-jump library results are available in
the supplementary (TODO:TABLE). Storing the uncompressed quality
values requires 1 byte per base-call because they are stored in
\textsc{ascii} format and is denoted by the dotted black line in the
figure.  The lossless compression of each data set using BZip2 ranges
from 2.19 - 3.10 bits/base-call and is denoted by the colored dotted
lines.  The compression methods tend to cluster together across the
different data sets.  Across all data sets, the 0-order polynomial
regression, profile encodings, and QualComp have the lowest
bits/base-call.

@@

As the order of the polynomial increases, the bits/base-call increase
and the \textsc{mse} decreases at an exponential rate.  The 7th-order
polynomial regression has the highest bits/base-call and in the
\textit{Mus musculus} data set, and requires more storage than the
\textsc{ascii} original quality values.  A 7th-degree polynomial
requires storing eight floating point values, resulting in 32 bytes
per quality sequence.  The read length of the \textit{Mus musculus}
data set is only 26, so the 7th-degree polynomial regression is storing
six more bytes than necessary for lossless encoding the quality data.

@@

No assembly of the \textit{Rhodobacter sphaeroides} 2.4.1 data set outperforms all others in all metrics (Table \ref{fig:assembly_ranks}).
%The assembly of the \textit{Rhodobacter sphaeroides} 2.4.1 data set using the uncompressed reads outperforms all compression methods in terms of \textsc{lap}, NG50, relocations, translations, and inversions.
Among the compression methods, the
256-, 64-, 128-profile encoding had the highest ranks, followed by QualComp with rates 10 bits/read, 30 bits/read, 100 bits/read, then 7th-order polynomial regression, followed by the QualComp with rate 6 bits/read, the 2-bin encoding, and lastly, the 3rd-order, 5th-order, and 0-order polynomials.

The lossy compression methods largely preserve the contiguity found in
the assembly produced using the reads with unmodified quality
sequences. All compression methods other than 0-order polynomial
regression produce an N50 ranging from 3.17--3.22 Mbps (see Supplemental Table TODO). Despite the
similar contiguity statistics, the different compression methods vary
noticeably in the amount of \textsc{snp}s. The order of polynomial has
an inverse relationship with the amount of \textsc{snp}s detected. The
2-bin and profile methods detected the least amount of \textsc{snp}s
compared to the reference genome, outperforming the assembly using the
original quality sequences. A more in-depth evaluation is needed to
determine whether these compression methods are missing actual
\textsc{snp}s.

@@

In general, the greater the polynomial order, the better overall assembly; however, the 5th-order polynomial regression performs slightly worse than the 3rd-order polynomial.
Their respective ranks in terms of N50 and relocations are fairly distant, which lowers the 5th-order polynomial's overall ranking barely below the 3rd-order's.
The 1st- and 0-order polynomial regression methods perform poor in all metrics except assembled bases.
One explanation is that the high error portions of sequences are being marked as high quality, so \textsc{allpaths-lg} is unable to trim or error correct the sequences.  Assembled sequences that overlap maybe unable to align to the errors at the end of the reads, artificially inflating the assembled genome size.

@@

2-bin encoding has the nice property of being simple to compute and
has good bits/base-call values. The 2-bin encoding suffers from having
a high \textsc{mse}, but fortunately, we have shown that in the case
of genome assembly, 2-bin encoding outperforms all polynomial
regressions encodings with degree less than 3. 2-bin encoding of the
fragment and short-jump libraries of \textit{Rhodobacter sphaeroides}
have \textsc{mse}s of $2.42\times$ and $10.76\times$ the 3rd-order
polynomial regression encodings, respectively. This further highlights
the importance of using additional contextual information of the
subsequent analyses when evaluating compressed quality values.

>>>>

We compare the \textsc{mse} versus bits/base-call of the
\textit{Rhodobacter sphaeroides} 2.4.1, \textit{Homo sapiens}
chromosome 14, \textit{Escherichia coli} str. K-12 MG1655, and
\textit{Mus musculus} data sets (Figure \ref{fig:mse_vs_bpbp}). We
only include the fragment libraries for the \textit{Rhodobacter
  sphaeroides} 2.4.1, and \textit{Homo sapiens} chromosome 14 data
sets, but the additional short-jump library results are available in
the supplementary (TODO:TABLE). Storing the uncompressed quality
values requires 1 byte per base-call because they are stored in
\textsc{ascii} format and is denoted by the dotted black line in the
figure.  The lossless compression of each data set using BZip2 ranges
from 2.19 - 3.10 bits/base-call and is denoted by the colored dotted
lines.  The compression methods tend to cluster together across the
different data sets.  Across all data sets, the 0-degree polynomial
regression, profile encodings, and QualComp have the lowest
bits/base-call.

@@

As the order of the polynomial increases, the bits/base-call increase
and the \textsc{mse} decreases at an exponential rate.  The 7th-degree
polynomial regression has the highest bits/base-call and in the
\textit{Mus musculus} data set, and requires more storage than the
\textsc{ascii} original quality values.  A 7th-degree polynomial
requires storing eight floating point values, resulting in 32 bytes
per quality sequence.  The read length of the \textit{Mus musculus}
data set is only 26, so the 7th-degree polynomial regression is storing
six more bytes than necessary for lossless encoding the quality data.

@@

No assembly of the \textit{Rhodobacter sphaeroides} 2.4.1 data set outperforms all others in all metrics (Table \ref{fig:assembly_ranks}).
%The assembly of the \textit{Rhodobacter sphaeroides} 2.4.1 data set using the uncompressed reads outperforms all compression methods in terms of \textsc{lap}, NG50, relocations, translations, and inversions.
Among the compression methods, the
256-, 64-, 128-profile encoding had the highest ranks, followed by QualComp with rates 10 bits/read, 30 bits/read, 100 bits/read, then 7th-degree polynomial regression, followed by the QualComp with rate 6 bits/read, the 2-bin encoding, and lastly, the 3rd-degree, 5th-degree, and 0-degree polynomials.

The lossy compression methods largely preserve the contiguity found in
the assembly produced using the reads with unmodified quality
sequences. All compression methods other than 0-degree polynomial
regression produce an N50 ranging from 3.17--3.22 Mbps (see Supplemental Table TODO). Despite the
similar contiguity statistics, the different compression methods vary
noticeably in the amount of \textsc{snp}s. The order of polynomial has
an inverse relationship with the amount of \textsc{snp}s detected. The
2-bin and profile methods detected the least amount of \textsc{snp}s
compared to the reference genome, outperforming the assembly using the
original quality sequences. A more in-depth evaluation is needed to
determine whether these compression methods are missing actual
\textsc{snp}s.

@@

In general, the greater the polynomial order, the better overall assembly; however, the 5th-degree polynomial regression performs slightly worse than the 3rd-degree polynomial.
Their respective ranks in terms of N50 and relocations are fairly distant, which lowers the 5th-degree polynomial's overall ranking barely below the 3rd-degree's.
The 1st- and 0-degree polynomial regression methods perform poor in all metrics except assembled bases.
One explanation is that the high error portions of sequences are being marked as high quality, so \textsc{allpaths-lg} is unable to trim or error correct the sequences.  Assembled sequences that overlap maybe unable to align to the errors at the end of the reads, artificially inflating the assembled genome size.

@@

2-bin encoding has the nice property of being simple to compute and
has good bits/base-call values. The 2-bin encoding suffers from having
a high \textsc{mse}, but fortunately, we have shown that in the case
of genome assembly, 2-bin encoding outperforms all polynomial
regressions encodings with degree less than 3. 2-bin encoding of the
fragment and short-jump libraries of \textit{Rhodobacter sphaeroides}
have \textsc{mse}s of $2.42\times$ and $10.76\times$ the 3rd-degree
polynomial regression encodings, respectively. This further highlights
the importance of using additional contextual information of the
subsequent analyses when evaluating compressed quality values.
